---
title: "inactivation: Software for modeling of microbial inactivation"
author: "Alberto Garre, Jose A. Egea, Pablo S. Fernandez"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{inactivation: Software for modeling of microbial inactivation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

__XX Hacer las gráficas en ggplot?? Eso implicaría añadir otra dependencia...__

## Introduction

R-package __inactivation__ implements useful functions for the modelization of microbial inactivation. It allows making predictions for some model parameters given as well as for adjusting models to experimental data collected in the laboratory. All the functions implemented in the __inactivation__ package are written in R.

Once installed, the __inactivation__ package can be loaded by writing,

```{r}
library(microinactivation)
```

__inactivation__ uses the packages `deSolve` (Soetaert, Petzoldt, and Setzer 2010) and `FME` (Soetaert, Petzoldt, and Setzer 2010) to both predict results and adjust common inactivation models to experimental data. The inactivation models most commonly used in the industry have been implemented in this package:

* Bigelow model __XXreference__
* Weibull-Mafart model (Mafart et al, 2002),
* Weibull-Peleg model __XXreference__
* Geeraerd model (Geeraerd et al, 2000).

The __inactivation__ package provides several functions which help in the modelization of microbial inactivation process:

* `predict_inactivation()` returns the predicted inactivation process, given a set of model parameters.

* `fit_isothermal_inactivation()` adjusts the parameters of a selected inactivation model to a set of isothermal experimental data.

* `fit_dynamic_inactivation()` fits the parameters of a given inactivation model to a set of dynamic inactivation experiments.

Moreover, several functions with miscellaneous information about the models implemented are provided.

In this document, a survey of the functionality of the __inactivation__ package is provided. This demonstration is based on two example data sets included in the package itself. These examples imitate a collection of experimental data obtained for a set of isothermal and dynamic experiments.

## Example data sets

The __inactivation__ package contains two data sets that will be used within this vignette to illustrate its capabilities.

### The example data set for isothermal fit

The __inactivation__ package includes the `isothermal_inactivation` data set, which imitates the data obtained during a set of isothermal inactivation experiments. This data set can be loaded as:

```{r}
data(isothermal_inactivation)
```

This example data set has 3 variables and 36 observations.

* The first column (`time`) provides the time at which the measurement was taken,

* the second column (`temp`) gives the temperature of the experiment,

* the logarithmic difference is given by the third column (`log_diff`).

```{r}
head(isothermal_inactivation)

plot(log_diff  ~ time, data=isothermal_inactivation, col = as.factor(temp))
title("Example dataset: isothermal_inactivation")
legend("topright", c("95ºC", "105ºC"), col = c(1, 2), pch = c(1,1), cex=0.5)
```

### The example data set for dynamic fit

__XX Añadir la temperatura de cada observación al data set??__

The `dynamic_inactivation` data set contained within the __inactivation__ package provides a `data.set` imitating the data obtained during a non-isothermal inactivation experiment. This data set can be loaded as follows:

```{r}
data(dynamic_inactivation)
```

This example data set contains 2 variables and 19 observations. The first column, `time`, reflects the time point at which the the observation was taken. The second one, `N` provides the number of microorganisms counted on the observation.

```{r}
head(dynamic_inactivation)
plot(log10(N)  ~ time, data=dynamic_inactivation)
title("Example dataset: dynamic_inactivation")
```

__XX Cambiar nombre del data set__

## Inactivation models implemented

The most commonly used models for the modelization of microbial inactivation have been implemented in the __inactivation__ package: Bigelow, Weibull-Mafart, Weibull-Peleg and Geeraerd. A brief mathematical description of each one of them is presented through this section.

### Bigelow's model for isothermal data

Bigelow's model __XXreference__ assumes that the inactivation process follows a first order kinetics and, therefore, a negative linear relation exists between the logarithm of the number of microorganisms ($N$) and the time ($t$).

$$ \mathrm{log}_{10}(N) = -\frac{1}{D_T(T)}t $$

The parameter $D_T$ (also named D-value) equals the inverse of the slope of the line. Its relationship with temperature ($T$) is described by the equation

$$ \mathrm{log}_{10}D_T(T) = \mathrm{log}_{10}D_R + \frac{T_R-T}{z} $$

where $T_R$ is a reference temperature and $D_R$ the D-value at this temperature. The parameter $z$ is usually named z-value and provides an indication of the sensitivity of the microorganism to temperature variations.

### Bigelow's model for nonisothermal data

Bigelow's model for nonisothermal data is developed from the isothermal one by calculating the first derivative of its constitutive equation with respect to time.

$$ \frac{d\mathrm{log}_{10}(N)}{dt} = - \frac{1}{D_T(T)} $$

Please note that for nonisothermal experiments, the temperature is time dependent. Therefore, $D_T(T)$ is also time dependent. This dependency is not taken into account in the derivation of the equation.

In this model, the evolution of $D_T(T)$ with temperatures is ruled by the same expression as for the isothermal model:

$$ \mathrm{log}_{10}D_T(T) = \mathrm{log}_{10}D_R + \frac{T_R-T}{z} $$

### Weibullian models

Weibullian models assume that the the inactivation process can be viewed as a failure phenomenon. According to this hypothesis, the time that each microorganism within the population can sustain the environmental conditions follows a Weibull-type probability distribution. Two models of this family have been implemented: Weibull-Mafart and Weibull-Peleg.

#### Weibull-Mafart model for isothermal inactivation

The Weibullian model presented by Mafart et al. (2002) for isothermal cases is described by the equation:

$$ \mathrm{log}_{10}S(t) = - \Big( \frac{t}{\delta(T)} \Big)^\beta $$

where $S = N/N_0$ represents the fraction of survivors and $t$ the time. $\delta$ and $\beta$ are respectively the rate and shape factors of the Weibull distribution.

In this model, $\beta$ is assumed constant and $\delta$ follows and exponential relation with temperature ($T$):

$$ \delta(T) = \frac {\delta_{ref}} {10^ {\Big( \frac{T-T_{ref}}{z} \Big)} } $$

where $T_{ref}$ is a reference temperature, $\delta_{ref}$ is the value of $\delta$ at this reference temperature and $z$ describes the sensitivity of the microorganism to temperature variations.

#### Weibull-Mafart model for nonisothermal inactivation

The dynamic Weibull-Mafart model is developed from the isothermal one following a process analogous to the one used for the Bigelow model. The rate of reduction at each time is calculated as the first derivative of the constitutive equation with respect to time:

$$ \frac{d\mathrm{log}_{10}S(t)}{dt} =  -\beta \frac{t^{\beta-1}}{\delta(T)^\beta} $$

Note that, as well as for Bigelow's model, the dependency of $\delta(T)$ with temperature has not been considered in the derivation.

As well as for the isothermal model, an exponential relation between $\delta(T)$ and temperature is assumed.

$$ \delta(T) = \frac {\delta_{ref}} {10^ {\Big( \frac{T-T_{ref}}{z} \Big)} } $$

#### Weibull-Peleg model for isothermal inactivation

The Weibullian model presented by __XXreference__ for isothermal cases follows the equation:

$$ \mathrm{log}_{10}S(t) = - b \cdot t^n $$

where $S = N/N_0$ represents the fraction of survivors and $t$ the time, and $b$ and $n$ are the two parameters of the model. Both parameters are equivalent to $\delta$ and $\beta$ parameters of the Weibull-Mafart model as they define the Weibull distribution function:

$$n = \beta$$
$$b = \frac {1}{\delta^\beta} $$

__XX a lo mejor quitar la comparación entre ambos modelos__

In the Weibull-Peleg model the parameter $n$ is assumed temperature independent, while the dependency of $b$ with temperature is modeled as

$$ b(T) = \mathrm{ln} \big\{ 1 + \mathrm{exp} \big[ k(T - T_c) \big] \big\} $$

This equation states that for temperatures much lower than the critical one ($T_c$) $b(T) = 0$, so there is no inactivation. For temperatures close to $T_c$, there is and exponential growth of $b(T)$ with respect to the temperature. For temperatures much higher than the critical one, a linear relation exists between $T$ and $b(T)$ with a slope $k$.

```{r}
Temperature <- seq(30, 80, length=100)
b <- log( 1 + exp(0.3*(Temperature - 60)))
b_1 <- log( 1 + exp(0.2*(Temperature - 60)))
plot(Temperature, b, type = "l", col="red")
lines(Temperature, b_1, type = "l", col = "blue")
legend("topleft", c("k = 0.3","k = 0.2"), col=c("red", "blue"), lwd=1, cex = 0.5)
```

#### Weibull-Peleg model for non-isothermal inactivation

The model developed for nonisothermal inactivation by Peleg in __XX reference__ has been implemented in the __inactivation__ package. This model begins with the hypothesis that isothermal survival curves are described by the equation:

$$ \mathrm{log}_{10}S(t) = -b(T) \cdot t^n $$

The inactivation rate for a constant temperature can be calculated as the first derivative with respect to time of this equation asuming that the model parameters do not depend on temperature.

$$ \Bigg| \frac{\mathrm{log}_{10}S(t)}{dt} \Bigg|_{T=const} = -b(T) \cdot n \cdot t^{n-1} $$

The time required to achieve any survival ratio, $t^\star$, can be calculated from the model equation as:

$$ t^\star = \Bigg| \frac{-\mathrm{log}_{10}S(t)}{b(T)} \Bigg|^{1/n} $$

Substituting this expression into the rate equation, the differential constitutive equation of the model is obtained:

$$ \Bigg| \frac{\mathrm{log}_{10}S(t)}{dt} \Bigg| =
    -b(T) \cdot n \cdot \Bigg| \frac{ -\mathrm{log}_{10}S(t) }{b(T)} \Bigg|^\frac{n-1}{n} $$

where $b(T)$ follows the same expression as for the isothermal model.

$$ b(T) = \mathrm{ln} \big\{ 1 + \mathrm{exp} \big[ k(T - T_c) \big] \big\} $$

### Geeraerd's model

The inactivation model described by Geeraerd et al. (2000) is based on the assumption that the inactivation process follows a first order kinetics with rate $k$. 

$$ \frac {dN}{dt} = - k \cdot N $$ 

However, the parameter $k$ is corrected with two parameters to include for shoulder ($\alpha$) and tail ($\gamma$) effects:

$$ k = k_{max} \cdot \alpha \cdot \gamma $$

where $k_{max}$ stands for the maximum inactivation rate of the microorganism at the current environmental conditions. A model equivalent to Bigelow's is used:

$$ k_{max} = \frac{\mathrm{ln}10}{D_{ref}} \cdot \mathrm{exp}
        \Big( \mathrm{ln}(10) \frac{T - T_{ref}}{z} \Big) $$

The shoulder parameter is defined as

$$ \alpha = \frac {1}{1+C_c} $$

where $C$ is a dummy component that limits the microbial inactivation. It is assumed that this component also follows a first order kinetics with rate $k_{max}$:

$$ \frac {dC_c}{dt} = - k_{max} \cdot C_c $$

The tail parameter is defined by the equation

$$ \gamma = 1 - \frac{N_{res}}{N} $$

where $N_{res}$ stands for the residual number of microorganism after the treatment.

## Prediction of microbial inactivation

The __inactivation__ package can predict the inactivation process of a microorganism through the `predict_inactivation()` function, which solves the differential equation required using the `lsoda` integrator.

The `predict_inactivation()` function requires four input variables:

* `simulation_model`
* `times`
* `parms`
* `temp_profile`

`simulation_model` is a string identifying the model to use. A list of the availables models can be obtained by calling the function `get_model_data()` without any argument.

```{r}
get_model_data()
```

In this example Geeraerd's model will be used, so the value `Geeraerd` will be assigned.

```{r}
example_model <- "Geeraerd"
```

`times` is a numeric vector which specifies at which values of time the results will be output. It does not define the points at which the numerical integration is done, so changing it will not affect the values calculated at individual time points.

```{r}
times <- seq(0, 5, length=100)
```

`parms` is a named numeric vector specifying the values of the model parameters, as well as the initial values for the integration. A call to the function `get_model_data()` with a valid model identifier as an input argument provides several data concerning this model as a list. Namely, its parameters and its variables.

```{r}
model_data <- get_model_data(example_model)
print(model_data$parameters)
print(model_data$variable)
```

The input arguments `parms` must provide values for all the model parameters, as well as initial values for all the variables of the model. The names of the model parameters must be identical to those provided by the function `get_model_data()`. On the other hand, the names assigned to the initial values of the values must be equal to the ones provided by `get_model_data()` plus a character \code{"0"}.

```{r}
model_parms <- c(D_R = 1,
                 z = 10,
                 N_min = 100,
                 temp_ref = 100,
                 N0 = 100000,
                 C_c0 = 1000
                 )
```

The last input variable required for the function is a data frame defining a discrete temperature profile. Values of the temperature at time points not provided will be approximated by using linear interpolation. A trapezoidal temperature profile will be used in this example.

```{r}
temperature_profile <- data.frame(time = c(0, 1.25, 2.25, 4.6),
                                  temperature = c(70, 105, 105, 70))
```

The prediction is calculated by calling the function `predict_inactivation()` with the aforementioned parameters.

```{r}
prediction_results <- predict_inactivation(example_model, times, model_parms, temperature_profile)
```

The value returned by the function is a list of class `SimulInactivation`. This list has four entries which provide enough information to reproduce the calculation:

* model: a string with the identifier of the model used for the simulation.
* model_parameters: a named numeric list with the model parameters used.
* temp_approximations: a list with two entries, each providing the functions used to approximate the temperature at each time point.
* simulation: a data frame with the results of the simulation

The entry _simulation_ provides a data frame with the calculated results.

```{r}
head(prediction_results$simulation)
```

The first column contains the times at which results are output are given. They are equal to the values provided by the `times` input variable.

The following columns provide the results obtained for the variables of the models. In this case, `N` and `C_c`. The last three columns provide the values of $\mathrm{log}_{10}(N)$, $S$ and $\mathrm{log}_{10}(S)$.

The `SimulInactivation` object includes S3 methods for the plotting of its results. Calling the function `plot()` with the object as arguments produces this plot.

```{r}
plot(prediction_results)
```

## Fitting of isothermal data

The package __inactivation__ can fit the parameters of an inactivation model to a set of isothermal experimental data through the function `fit_isothermal_inactivation()`. This function makes use of the `nls` function included in the `stats` package to fit the parameters of an inactivation model to a set of experimental data.

The experimental data to fit is provided by the data frame variable `death_data`. It must contain at least three columns providing the times at which the observations were made (_time_), the temperature of the experiment (_temp_) and the logarithmic difference observed with respect to the orginal population (*log_diff*). The example data set `isothermal_inactivation` will be used for this demonstration.

```{r}
data(isothermal_inactivation)
```

The model to use for the adjustment is defined by the character variable `model_name`. A list of the valid model keys for isothermal adjustment can be obtained by calling the function `get_isothermal_model_data()` without any arguments.

```{r}
get_isothermal_model_data()
```

Bigelow's model, whose key is `"Bigelow"`, will be used in this example.

```{r}
model_name <- "Bigelow"
```

The names of the parameters of an inactivation model can be obtained by calling the function `get_isothermal_model_data()` with the model key as argument.

```{r}
model_data <- get_isothermal_model_data(model_name)
model_data$params
```

Therefore, the isothermal Bigelow model requires the definition of three parameters: `z`, `D_ref` and `ref_temp`. The `fit_isothermal_inactivation()` function allows to distinguish between known and unknown (i.e., to be adjusted) model parameters. The known model parameters are set by the input argument `known_parameters`, which must be a `list`. The reference temperature is usually set, so we will fix it to 100ºC in this example.

```{r}
known_params = list(temp_ref = 100)
```

Due to the use of the `nls` functions, initial points are required for the unknown model parameters. These initial values are provided by the input argument `starting_point`. This variable needs to be a named numeric vector.

```{r}
starting_point <- c(z = 10,
                    D_R = 1
                    )
```

The `fit_isothermal_inactivation()` functions allows to make the adjustment based on the minimization of the error of either the logarithmic count or the total number of microorganisms. In this example, the adjustment will be based on the minimization of the error of the logarithmic count, so the `adjust_log` argument will be set to `TRUE`.

```{r}
adjust_log <- TRUE
```

The adjustment is made by calling the `fit_isothermal_inactivation()` function with the arguments which have just been defined. 

```{r}
iso_fit <- fit_isothermal_inactivation(model_name, isothermal_inactivation,  
                                       starting_point, adjust_log, known_params)
```
This function returns a list of class `IsoFitInactivation`. This object contains the results of the adjustment, as well as the statistical information of the adjustment process. This information is divided in four entries:

* The information of the adjustment process is provided under the entry `nls`, where the object returned by the `nls` function is saved.

```{r}
iso_fit$nls
```

* The values of the parameters fixed under the input argument `known_arguments` as well as those calculated by the adjustment are provided as a list. This list can be accessed under the entry `parameters`.

```{r}
iso_fit$parameters
```

* The key of the model used for the adjustment is provided under the header `model`.

```{r}
iso_fit$model
```

* The data used for the adjustment is saved  under the header `data`.

```{r}
head(iso_fit$data)
```

The class `IsoFitInactivation` implements S3 methods for `plot`, which allow the creation of a comparison plot between the best prediction and the experimental results.

```{r}
plot(iso_fit)
```

## Fitting of dynamic experiments

The __inactivation__ package implements the function `fit_dynamic_inactivation()`, which allows the adjustment of inactivation models to experimental data with time dependent temperature profiles.

The function `fit_dynamic_inactivation()` requires several input arguments:

* The experimental data to use for the fit.
* A string identifying the inactivation model to use.
* The temperature profile of the experiment.
* The values of the known model parameters, as well as starting points and bounds for the unknown ones.
* A logical specifying whether the adjustment will be based on the error of the logarithmic count or the total count.

The experimental data to fit is provided by the variable `experiment_data`. It has to be a data frame with at least a column named _time_, which provides the time at which the measurements were taken, and another one named _N_, with the number of microorganisms counted at each time. The data set `dynamic_inactivation` included in the __inactivation__ package will be used for this demonstration.

```{r}
data(dynamic_inactivation)
```

The model to use for the fitting is defined by the character variable `simulation_model`. A list of the available models for dynamic adjustment can be obtained by calling the function `get_model_data()` without any input argument.

```{r}
get_model_data()
```

The Weibull-Peleg model, whose key is *Weibull_Peleg* will be used in this example.Therefore, we assign it to variable `simulation_model`.

```{r}
simulation_model <- "Weibull_Peleg"
```

The temperature profile of the experiment is given by the data frame `temp_profile`. This variable provides the values of the temperature at discrete time points. Temperatures at time points not provided are calculated by linear interpolation. For this example, a generic trapezoidal temperature profile will be used:

```{r}
dummy_temp <- data.frame(time = c(0, 1.25, 2.25, 4.6),
                         temperature = c(70, 105, 105, 70))
```

The `fit_dynamic_inactivation()` functions permits the distinction between known and unknown (i.e., to be adjusted) model parameters. The known model parameters are asigned to the argument `fixed_parameters`, which is a named numeric vector.

The names of model parameters can be retrieved by calling the function `get_model_data()` with the model key as argument.

```{r}
model_data <- get_model_data(simulation_model)
model_data$parameters
```

The parameter *temp_crit* will be considered known equal to 100ºC in this example, so we will assign it to the `fixed_parameters` argument.

```{r}
fixed_parameters = c(temp_crit = 100)
```

The rest of the model parameters, as well as number of microorganism at time zero, will be adjusted to the experimental data. The `fit_dynamic_inactivation()` function makes the adjustment through the function `modFit` of the `FME` package (Soetaert et al. 2010). This function requires the definition of both starting points and bounds for every single one of the parameters to adjust. This information is passed by the input arguments `starting_points`, `upper_bounds` and `lower_bounds`, which share format with the input argument `fixed_parameters`. In case of unbounded variables, `+Inf` or `-Inf` must be assigned to the desired entries of `upper_bounds` or `lower_bounds`.

```{r}
starting_points <- c(n = 1,
                     k_b = 0.25,
                     N0 = 1e+05)
```
Reasonable lower and upper bounds will be defined, so variables cannot be negative or unrealistically large.
```{r}
upper_bounds <- c(n = 2,
                  k_b = 1,
                  N0 = Inf)

lower_bounds <- c(n = 0,
                  k_b = 0,
                  N0 = 1e4)
```

Note that the initial number of microorganism (`N0`) has also been assigned initial values and bounds. This definition is analogous to the one of the model parameters, although its name must be equal to the one provided by the function `get_model_data()` plus a caracter `"0"` . In this example, the initial value of the variable `"N"` has been named `"N0"`.

Care must be taken when setting initial points and bounds, because if they are unreasonable the fitting algorithm will fail.

The adjustment can be based on the minimization of the error of the total count of microorganism ($N$) or the logarithmic count ($\mathrm{log}_{10}N$). This distinction is made by the boolean input argument `minimize_log`. In case it is `TRUE`, the error of the logarithmic count is minimized. For this example, the error in the total number of microorganism will be optimized. We, thus, define the variable `minimize_log` accordingly:

```{r}
minimize_log = TRUE
```

The adjustment is made by calling the function `fit_dynamic_inactivation()` with the variables which have been just defined.

```{r}
dynamic_fit <- fit_dynamic_inactivation(dynamic_inactivation, simulation_model, dummy_temp,  
                                        starting_points, upper_bounds, lower_bounds,  
                                        fixed_parameters, minimize_log)
``` 

The function returs a list of class `FitInactivation` with three entries:

* *fit_results* provides the instance of `modFit` returned by the `FME::modFit()` function.

* under *best_prediction* an instance `SimulInactivation` with the results of the best prediction is saved.

* the header _data_ contains the experimental data used for the fit.

```{r}
names(dynamic_fit)
```

The parameters of the model calculated are saved under the entry `model_parameters` of the `SimulInactivation` instance.

```{r}
dynamic_fit$best_prediction$model_parameters
```

The `FitInactivation` object includes S3 methods for the creation of a comparison plot between the experimental results and the prediction made by the model parameters adjusted.

```{r}
plot(dynamic_fit)
```

## References

__XX Ordenar referencias__  
__XX Homogeneizar formatos referencias__

Karline Soetaert, Thomas Petzoldt, R. Woodrow Setzer (2010).  *Solving
Differential Equations in R: Package deSolve Journal of Statistical Software*,
33(9), 1--25.  URL http://www.jstatsoft.org/v33/i09/.

Soetaert, Karline and Petzoldt, Thomas, 2010. *Inverse Modelling, Sensitivity
and Monte Carlo Analysis in R Using Package FME. Journal of Statistical
Software*, 33(3), 1-28. URL http://www.jstatsoft.org/v33/i03/.

Mafart, P., Couvert, O., Gaillard, S., & Leguerinel, I. (2002). On calculating
sterility in thermal preservation methods: Application of the Weibull
frequency distribution model. International Journal of Food Microbiology, 72,
107–113.

Geeraerd A.H., C.H. Herremans and J.F. Van Impe 2000. “Structural model requirements to describe microbial
inactivation during a mild heat treatment.” International Journal of Food Microbiology, 59, 185-209.
